---
title: "Task W - Bivariate Correlations"
author: "Clara Khuon"
date: "January 15th, 2026"
output: html_document
---

## Correlations for Task W with PHQ Scores at Timepoint 3 ##

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Set working directory
setwd("...")

# Load libraries
library(tidyverse)
library(here)
library(caret)
library(lattice)
library(parallel)
library(doParallel)
library(foreach)
library(glmnet)
library(Hmisc)
library(randomForest)
library(lsr)
library(car)
library(ggplot2)
library(corrplot)

# Load data
demographic_data <- read_csv(file = here("demographic_data.csv"))
taskW_TAASSC <- read_csv(file = here("taskW_linguistic_data_TAASSC.csv"))
taskW_SCA <- read_csv(file = here("taskW_linguistic_data_SCA.csv"))

# Note: File names above are example placeholders

```

# Preprocessing 1
```{r}
# Filter demographic dataset to specific variables
taskW_demographic_variables <- c("UserID_v2", "T3_PHQ9_Total", "Age", "Gender", "Education", "Diagnosis_Onset", "TaskW_1_Acceptability1", "TaskW_1_Acceptability2", "TaskW_1_Acceptability3")
taskW_demographic_data <- demographic_data[, taskW_demographic_variables]

# Order SCA dataset in ascending order of participant ID number
taskW_SCA$filename <- basename(taskW_TAASSC$filename)
taskW_SCA$filename <- str_remove(taskW_SCA$filename,".txt")
taskW_SCA$filename <- str_remove(taskW_SCA$filename,"TaskW.Participant")
taskW_SCA <- taskW_SCA[order(as.numeric(taskW_SCA$filename), decreasing = FALSE), ]
taskW_SCA$filename <- paste("Participant.", taskW_SCA$filename, sep = "")

# Order TAASSC dataset in ascending order of participant ID number
taskW_TAASSC$filename <- str_remove(taskW_TAASSC$filename,".txt")
taskW_TAASSC$filename <- str_remove(taskW_TAASSC$filename,"TaskW.Participant")
taskW_TAASSC <- taskW_TAASSC[order(as.numeric(taskW_TAASSC$filename), decreasing = FALSE), ]
taskW_TAASSC$filename <- paste("Participant.", taskW_TAASSC$filename, sep = "")

# Remove predictor variables with near-zero variance from TAASSC dataset
near_zero_table <- nearZeroVar(taskW_TAASSC, saveMetrics = TRUE)
near_zero_TAASSC <- nearZeroVar(taskW_TAASSC) 
taskW_TAASSC <- taskW_TAASSC[,-near_zero_TAASSC]

# Combine demographic dataset with TAASCC and SCA datasets (without duplicate participant ID number and word-count columns)
taskW_total <- cbind(taskW_demographic_data, taskW_TAASSC, taskW_SCA[,3:16])

# Identify and remove participants with invalid writing samples (i.e., word-count of zero)
invalid <- filter(taskW_total, nwords == 0)
taskW_filtered <- filter(taskW_total, nwords != 0)

# Remove second ID column
taskW_filtered <- taskW_filtered[,-which(colnames(taskW_filtered) == "filename")]

# Define gender and education as categorical variables in final dataset
taskW_filtered$Gender <- as.factor(taskW_filtered$Gender)
taskW_filtered$Education <- as.factor(taskW_filtered$Education)

# Final dataset
taskW_final <- taskW_filtered
taskW_final

```

# Descriptive Statistics of Writing Samples
```{r}
# Average number of words across writing samples
mean(taskW_final$nwords)

# SD number of words across writing samples
sd(taskW_final$nwords)

# Min number of words across writing samples
min(taskW_final$nwords)

# Max number of words across writing samples
max(taskW_final$nwords)

```

# Descriptive Statistics of Acceptability Scores of Final Participant Sample
```{r}

# Acceptability 1 = Difficulty
length(taskW_final$TaskW_1_Acceptability1)
mean(taskW_final$TaskW_1_Acceptability1)
sd(taskW_final$TaskW_1_Acceptability1)
min(taskW_final$TaskW_1_Acceptability1)
max(taskW_final$TaskW_1_Acceptability1)

# Acceptability 2 = Interest
length(taskW_final$TaskW_1_Acceptability2)
mean(taskW_final$TaskW_1_Acceptability2)
sd(taskW_final$TaskW_1_Acceptability2)
min(taskW_final$TaskW_1_Acceptability2)
max(taskW_final$TaskW_1_Acceptability2)

# Acceptability 3 = Willingness
length(taskW_final$TaskW_1_Acceptability3)
mean(taskW_final$TaskW_1_Acceptability3)
sd(taskW_final$TaskW_1_Acceptability3)
min(taskW_final$TaskW_1_Acceptability3)
max(taskW_final$TaskW_1_Acceptability3)

```

# Total PHQ-9 Score Descriptive Statistics
```{r}
# Descriptive Statistics
length(taskW_final$T3_PHQ9_Total) # total number of participants with PHQ-9 total data
mean(taskW_final$T3_PHQ9_Total)
sd(taskW_final$T3_PHQ9_Total)
min(taskW_final$T3_PHQ9_Total)
max(taskW_final$T3_PHQ9_Total)

# Total Number by Symptom Severity Category
nil_to_minimal <- c(0, 1, 2, 3, 4)
mild <- c(5, 6, 7, 8, 9)
moderate <- c(10, 11, 12, 13, 14)
moderately_severe <- c(15, 16, 17, 18, 19)
severe <- c(20, 21, 22, 23, 24, 25, 26, 27)

nrow(filter(taskW_final, T3_PHQ9_Total %in% nil_to_minimal)) # Nil to Minimal
nrow(filter(taskW_final, T3_PHQ9_Total %in% mild)) # Mild
nrow(filter(taskW_final, T3_PHQ9_Total %in% moderate)) # Moderate
nrow(filter(taskW_final, T3_PHQ9_Total %in% moderately_severe)) # Moderately Severe
nrow(filter(taskW_final, T3_PHQ9_Total %in% severe)) # Severe

# Total PHQ-9 Score >= 10 (Probable Depression)
nrow(filter(taskW_final, T3_PHQ9_Total >= 10))

```

# Preprocessing 2
Identify and remove highly-correlated predictor variables
```{r}
# Isolate dataset columns with predictor variables
taskW_variables <- taskW_final[,which(colnames(taskW_final)=="nwords"):dim(taskW_final)[2]]

# Create correlation matrix of predictor variables
cor_matrix <- cor(taskW_variables, use = "complete.obs")

# Identify highly correlated variables
highly_correlated <- findCorrelation(cor_matrix, cutoff = 0.75)

# Add number of columns to vector of column names to align with length of original filtered dataset
highly_correlated <- highly_correlated + length(taskW_demographic_variables)

# Remove highly correlated variables from original filtered dataset
taskW_final <- taskW_final[,-highly_correlated]

# Final dataset
taskW_final 

```

# Preprocessing 3
Correlations between linguistic variables and total PHQ-9 scores
With FDR correction

**NOTE: If applying edits that change dimensions of dataset, column reference numbers below need to be updated**

```{r}
taskW_final_df <- as.data.frame(taskW_final)
taskW_pvalues <- vector()

j = 1
for(i in 10:109){
  taskW_pvalues[j] <- cor.test(taskW_final_df$T3_PHQ9_Total, taskW_final_df[,i])$p.value
  j <- j+1
}

# Specify 10:109 so only runs over linguistic variables

taskW_pvalues <- p.adjust(taskW_pvalues, method = "fdr")
taskW_variables_refined <- colnames(taskW_final[,10:109])[which(taskW_pvalues < .05)]
taskW_final <- taskW_final[, c(taskW_demographic_variables,taskW_variables_refined)]
taskW_final 

# no variables are significantly correlated with PHQ total after applying FDR correction

```

## Significant Bivariate Correlations From All Tasks
```{r}

taskW_cor1 <- cor.test(taskW_final_df$T3_PHQ9_Total, taskW_final_df$nn_pobj_deps_NN_struct)
taskW_cor2 <- cor.test(taskW_final_df$T3_PHQ9_Total, taskW_final_df$fic_lemma_ttr) 
#taskW_cor3 <- cor.test(taskW_final_df$T3_PHQ9_Total, taskW_final_df$all_construction_ttr) 
#taskW_cor4 <- cor.test(taskW_final_df$T3_PHQ9_Total, taskW_final_df$all_lemma_construction_attested) 
#taskW_cor5 <- cor.test(taskW_final_df$T3_PHQ9_Total, taskW_final_df$news_av_lemma_construction_freq_type) 
#taskW_cor6 <- cor.test(taskW_final_df$T3_PHQ9_Total, taskW_final_df$news_av_lemma_construction_freq_log_stdev) 

# 4 of the variables had been removed during preprocessing for TaskW

```

# Create Results Dataframe
With FDR-corrected p-values
```{r}

# create empty 6 x 3 matrix; convert matrix to dataframe; rename column names
taskW_results <- matrix(NA, nrow = 6, ncol = 3)
taskW_results <- data.frame(taskW_results) 
colnames(taskW_results) <- c("variable", "r", "p_value_adjusted")       

# fill dataframe with values from correlation results

# noun phrase complexity
taskW_results[1,1] <- "nn_pobj_deps_NN_struct"                      # row1 col1 = variable name
taskW_results[1,2] <- taskW_cor1$estimate                           # row1 col2 = pearson's r from cor1
taskW_results[1,3] <- p.adjust(taskW_cor1$p.value, method = "fdr")  # row1 col3 = FDR adjusted p-value from cor1

taskW_results[2,1] <- "fic_lemma_ttr"    
taskW_results[2,2] <- taskW_cor2$estimate         
taskW_results[2,3] <- p.adjust(taskW_cor2$p.value, method = "fdr")          

taskW_results[3,1] <- "all_construction_ttr"    
taskW_results[3,2] <- "NA"         
taskW_results[3,3] <- "NA"          

taskW_results[4,1] <- "all_lemma_construction_attested"    
taskW_results[4,2] <- "NA"         
taskW_results[4,3] <- "NA"          

taskW_results[5,1] <- "news_av_lemma_construction_freq_type"    
taskW_results[5,2] <- "NA"         
taskW_results[5,3] <- "NA"          

taskW_results[6,1] <- "news_av_lemma_construction_freq_log_stdev"    
taskW_results[6,2] <- "NA"         
taskW_results[6,3] <- "NA"

taskW_results

# NA = variables removed during preprocessing, but were significant in other tasks

```
